Title here

### 2020-07-29 paper/paper.Rmd
modify based on meeting 19 and polish section 3 and 4 (except 4.1)

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their intial paper on projection pursuit in 1974 that "the  technique  use  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied to more general to help practitioner to unveil the black-box in randomised iteartive (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

The paper is organised as follows. Section \ref{DFO} gives a general literature review of optimisation, specifically derivative free optimisation. Section \ref{vis-diag} presents the new idea of constructing a
systematic visual framework that diagnoses the components of an optimisation procedure (parameters, searching path, etc). The rest of the paper serves as a comprehensive example of using the visual diagnostics on one particular problem: *projection pursuit guided tour*. Some background knowledge of projection pursuit guided tour is provided in Section \ref{tour}. Section \ref{apply} applies the concepts proposed in section \ref{vis-diag} in the tour problem and sets up the data. The last section, Section \ref{plots}, presents the visual diagnostic plots and explains how they can help to understand different aspects of the optimisation in tour.

@@

From Section \ref{tour}, we presents a comprehensive case study on how to use the visual diagnostics to explore the optimisation in a specific problem: projection pursuit guided tour. Section \ref{tour} aims to provide non-experts with an overview of the problem content and the existing optimisation procedures used in projection pursuit guided tour. For those who are already familiar with the techniques, feel free to skip this section.

The optimisation problem we're interested in is in the context of projection pursuit. Coined by @friedman1974projection, projection pursuit is a method that detects the interesting structure (i.e. clustering, outliers and skewness) of multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be a data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and can be used to measure the "interestingness" of the projection. A number of index functions have been proposed in the literature including Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], natural Hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

 As @friedman1974projection noted "..., the technique use for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." A suitable optimisation procedure is needed to find the projection angle that maximises the PPI and the quality of the optimisation largely affect the interesting projections one could possibly observe.

Projection pursuit is usually used in conjunction with a tour method called *guided tour*. Tour explores the multivariate data *interactively* via playing a series of projections, that form a *tour path* and guided tour uses the path that is geodesically the shortest. Details of the mathematical construction of a tour path can be found in @buja2005computational.  [Maybe add the sketch for illustrating tour path.] Guided tour, along with other types of tour, has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

@@

```{r tour-path, out.width="100%", out.height="60%", fig.cap="\\label{tour-path}An illustration of the tour path"}
knitr::include_graphics(path = here::here("figures/tour_path_increasing.png"))
```

## Optimisation problem and existing algorithms

Now we begin to formulate the optimisation problem. Given a randomly generated starting basis $\mathbf{A}^0$, projection pursuit finds the final projection basis $\mathbf{A}^J = [\mathbf{a}_1, \cdots, \mathbf{a}_d]$, where $\mathbf{a}_i$ is a $p \times 1$ vector, satisfies the following optimisation problem:

@@

via an iterative search of target basis: $\{\mathbf{A}^0, \mathbf{A}^1, \mathbf{A}^2, \cdots \cdots \mathbf{A}^J\}$.

There are three existing methods for optimising PPI function and we review them below.

@posse1995projection proposed a stochastic direct search method, a random search algorithm. Given a current basis $\mathbf{A}_{\text{cur}}$, a candidate basis $\mathbf{A}_{t}$ is sampled in the neighbourhood defined by the radius of the p-dimensional sphere, $\alpha$ of the current basis $\mathbf{A}_{t}$ by $\mathbf{A}_{t} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$. If the candidate basis has a higher index value than the current basis, it is outputted as the target basis $\mathbf{A}_{t}$ along with other metadata and the current iteration stops. If no basis is found to have higher index value after the maximum number of tries $l_{\max}$, the algorithm stops with nothing outputted. $c$ is the halfing parameter and when $c > 30$, the neighbourhood parameter $\alpha$ in the next iteration will be reduced by half. The algorithm for a random search is summarised in Algorithm \ref{random-search}.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{t+1}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

@cook1995grand explained the use of a gradient ascent optimisation with the assumption that the index function is continuous and differentiable. Since some indices could be non-differentiable, the computation of derivative is replaced by a pseudo-derivative of evaluating five randomly generated directions in a tiny nearby neighbourhood. Taking a step on the straight derivative direction has been modified to maximise the projection pursuit index along the geodesic direction. See Algorithm \ref{search-geodesic} for a summarisation of the steps.

@@

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(t)$. Given an initial $T_0$, the temperature at iteration $t$ is defined as $T(t) = \frac{T_0}{log(t + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{rand}$, a candidate basis is constructed as $\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis becomes the target basis. If it is smaller, the candidate is accepted with probability $P= \min\left\{\exp\left[\frac{I_{t+1} - I_{t}}{T(t)}\right],1\right\}$ where $I_{t}$ and $I_{t+1}$ are the index value of the current and candidate basis respectively. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The global object; The target basis: $\mathbf{A}^{j}$}
  initialisation\;
  \While{$l < l_{\max}$}{
    Set $l = 0$\;
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

Below listed several features characterise the optimisation procedures needed in projection pursuit

- *Being able to handle non-differentiable PPI function*: The PPI function could be non-differentiable, thus derivative free methods are preferred.

- *Being able to optimise with constraints*: The constraint comes from projection matrix being an orthonormal matrix.

- *Being able to find both local and global maximum*: Although the primary interest is to find the global maximum, local maximum could also reveal structures in the data that are of our interest.

@@

Random search methods has a black-box mechanism and focuses solely on finding the global maximum point, while the projection pursuit problem we have aims at *exploring* the data and thus is interested in how the algorithm finds its maximum. This motivates us **to develop a visual diagnostic system for exploring the optimisation searching path**.

@@

Ggplot requires a data frame that contains all the variables to plot and a *global object* is constructed as the data frame supplied to the visual diagnostic plots to better suit the characters of iterative optimisation algorithms. Given an optimisation algorithm, two primary variables of interest are the *decision variable: $\mathbf{A}$* and the *value of the objective function: $I$*, which is the projection basis and index value repsectively in projection pursuit. *State: $S$* is another variable of great importance since it labels $\mathbf{A}$ and $I$ by the stage they are in the whole search-and-interpolate process.

*Iterators* indexes the data collected and prescribes the the natural ordering of each observation in different levels. *$t$* is a unique identifier that has an increment of one for each observation and *$j$* is the counter for each search-and-interpolate round. [add a bit more for *$l$* after decide whether to combine $l$ and $k$. ]

@@

There are other parameters that are of our interest and we denote them as *$V_{p}$*. In projection pursuit, this includes $V_1 = \text{method}$, denoting the name of the searching method used and $V_2 = \text{alpha}$, which is a neighbourhood parameter that controls the size in sampling candidate bases. A sketch of the global object for projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}. The full data structure can thus be shown as in Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|ccc|ccc|ccc}
t & \mathbf{A} & I & S & j &  l  & k & V_{1}& \cdots & V_{p}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & \ast & V_{1, 1} & \cdots & V_{p, 1}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1 & \ast & V_{1, 2} & \cdots & V_{p, 2}\\
3 & \mathbf{A}_3 & I_3 & S_3 & 2 & 2 & \ast & V_{1, 3} & \cdots & V_{p, 3}\\
\vdots & \vdots &\vdots &\vdots & \vdots &\vdots & \vdots &\vdots &\cdots &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_1 & \ast & \vdots &  \cdots & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2 & \ast & 1& \vdots &  \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & 2& \vdots& \cdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & \ast & k_1&\vdots & \cdots & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots  &\vdots & \vdots &\vdots &\cdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_J & \ast & V_{1, T}& \cdots & V_{p, T}\\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

@@

We first simulate some random variables of size 1000 with different structures. `x1`, `x8`, `x9` and `x10` are simulated from normal distribution with zero mean and variance of one as in equation \ref{eq:sim-norm}. When using projection pursuit to explore the data structure based on its departure from normality, the entry in the projection basis for these variables should be close to zero in theory. `x2` to `x7` are mixture of normal distribution with different weights and locations. Equation \ref{eq:sim-x2} to \ref{eq:sim-x7} show the distribution of each variables and Figure \ref{origin-data} shows the histogram of each variable except `x3`. All the variables are then scaled to ensure the mixture has variance of one.

@@

We form our first dataset using variable `x1`, `x2`, `x8`, `x9` and `x10` and supply it to one of the above algorithm, say `search_geodesic` in the `tourr` package. When the optimisation ends, the global object will be stored and printed (it can be turned off by supplying argument `print = FALSE`). Additional messages during the optimisation can be displayed by `verbose = TRUE`. Below shows the first five rows of the global object. The global object meets the definition of tidy data introduced in [@wickham2014tidy] where each observation forms a row, each variable forms a column and each type of observational unit forms a table. The adoption of tidy data makes it easier for data wrangling and visualisation, which powers the visual diagnostics system introduced in this paper.

@@

```{r glb-obj-1}
print(holes_1d_better %>% head(5))
```

@@

# Visual diagnostics

@@

- Layer 1: boxplot geom
  - data: group by $j$ and filter the observations in the group that have count greater than `cutoff = 15`.
  - x: $j$ is mapped to the x-axis
  - y: the statistical transformed index value: $Q_q(I_{T} \cdot \mathds{1} (S = \text{search}))$ is mapped to the y-axis where $q$ takes one of $0, 25, 50, 75, 100$ and $Q_q(x)$ finds the qth-quantile for the vector $x$. $\mathds{1}(.)$ is the identity operator.

@@

# Implementation: Ferrn package

>>>>

- name: Author 1
  thanks: The authors gratefully acknowledge ...
  affiliation: Department of YYY, University of XXX
- name: Author 2
  affiliation: Department of ZZZ, University of WWW
keywords:
- optimisation
- projection pursuit 
- guided tourr
- visual 
- diagnostics 
- R
abstract: Friedman & Tukey commented on their intial paper on projection pursuit in 1974 that "the  technique  use  for  maximising  the  projection index strongly influences both the statistical and the computational aspects of the procedure." While many projection pursuit indices have been proposed in the literature, few concerns the optimisation procedure. In this paper, we developed a system of diagnostics aiming to visually learn how the optimisation procedures find its way towards the optimum. This diagnostic system can be applied to more general to help practitioner to unveil the black-box in randomised iteartive (optimisation) algorithms. An R package, ferrn, has been created to implement this diagnostic system.
bibliography: biblio.bib
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::asa_article
editor_options: 
  chunk_output_type: console
---

@@

The remainder of the paper is organised as follows. Section \ref{DFO} gives a general overview of optimisation methods, specifically derivative free optimisation. 
Section \ref{tour} introduces the optimisation problem in a specific problem: projection pursuit guided tour and existing derivative free algorithms for solving it. 
Section \ref{vis-diag} and \ref{vis-diag-plots} together present the new construction of a visual diagnostic framework that unveil the blackbox of optimisation. Section \ref{vis-diag} focues on motivational ideas and data structure and Section \ref{vis-diag-plots} provides definitions of different diagnostic plots using grammar of graphic and examples. 
Finally, \ref{implementation} mentions the R package: ferrn that implements the visual diagnostics described above.

@@

The optimisation problem we're interested in is in the context of projection pursuit guided tour. Coined by @friedman1974projection, projection pursuit detects interesting structures (i.e. clustering, outliers and skewness) in multivariate data via projecting it in lower dimensions. Let $\mathbf{X}_{n \times p}$ be the data matrix, an n-d projection can be seen as a linear transformation $T: \mathbb{R}^p \mapsto \mathbb{R}^d$ defined by $\mathbf{P} = \mathbf{X} \cdot \mathbf{A}$, where $\mathbf{P}_{n \times d}$ is the projected data and $\mathbf{A}_{p\times d}$ is the projection basis. Define $f: \mathbb{R}^{p \times d} \mapsto \mathbb{R}$ be an index function that maps the projection basis $\mathbf{A}$ onto a scalar index value $I$, this function is commonly known as the projection pursuit index (PPI) function, or the index function and is used to measure the "interestingness" of the projection. A number of index functions have been proposed in the literature to detect different data structures, including Legendre index [@friedman1974projection], Hermite index [@hall1989polynomial], natural Hermite index [@cook1993projection], chi-square index [@posse1995projection], LDA index [@lee2005projection] and PDA index [@lee2010projection].

In their initial paper, @friedman1974projection noted that "..., the technique use for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." Hence, effective optimisation algorithms are necessary for projection pursuit to find bases that give interesting projections. While we leave the formal construction the optimisation problem and existing algorithms in section \ref{tour-optim}, we want to emphasizes here that given a random starting (current) basis, projection pursuit repetitively searches for candidate bases nearby and outputs a target basis once it finds one with higher index value than the current basis. The target basis becomes the current basis in the next iteration and the searching continues until no better basis can be found or the termination criteria are reached.

A tour method, guided tour is usually used in conjunction with projection pursuit to interactively explore multivariate data. Tour plays an animation on the projected data which is constructed using the bases outputted by projection pursuit and their interpolations. Guided tour constructs the interpolation using the shortest geodesic path between two outputted bases and we refer readers to @buja2005computational for details of the mathematical construction of the interpolation basis  Figure \ref{tour-path} shows a sketch of the tour path consisting of the blue frames produced by projection pursuit and the white frames, which interpolates two blue frames. The tour method has been implemented in the *tourr* package in R, available on the Comprehensive R Archive Network at [https://cran.r-project.org/web/packages/tourr/](https://cran.r-project.org/web/packages/tourr/) [@wickham2011tourrpackage].

@@

```{r tour-path, out.width="100%", out.height="60%", fig.cap="\\label{tour-path}An illustration of the tour path"}
knitr::include_graphics(path = here::here("figures/tour_path_keynote/tour_path_keynote.001.jpeg"))
```

## Optimisation problem and existing algorithms {#tour-optim}

Now we begin to formulate the optimisation problem. Given a randomly generated starting basis $\mathbf{A}_1$, projection pursuit finds the final projection basis $\mathbf{A}_T = [\mathbf{a}_1, \cdots, \mathbf{a}_d]$, where $\mathbf{a}_i \in \mathbb{R}^{p}$, satisfies the following optimisation problem:

@@

There are three existing algorithms.

@posse1995projection proposed a random search algorithm that samples a candidate basis $\mathbf{A}_{l}$ in the neighbourhood of the current basis $\mathbf{A}_{\text{cur}}$ by $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$, where $\alpha$ controls the radius of the sampling neighbourhood and $\mathbf{A}_{rand}$ is a randomly generated matrix with the same dimension as $\mathbf{A}_{\text{cur}}$. The optimiser keeps sampling bases near the current basis until it finds one with higher index value than the current basis and then outputs it for guided tour to construct the interpolation path. A new round of search continues to find a better basis after the interpolation finishes. The halfing parameter $c$ with default value of 30 is designed to adjust the searching neighbourhood $\alpha$. When the search needs to sample more than $c$ number of basis to find an outputted basis, the neighbourhood parameter $\alpha$ will be reduced by half in the next iteration. If the optimiser can't find a better basis within the maximum number of tries $l_{\max}$, the algorithm stops. The algorithm is summarised in Algorithm \ref{random-search} for one iteration.

\begin{algorithm}
\SetAlgoLined
  \SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$}
    \Output{The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$ and $c = 0$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = \mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{\text{rand}}$ ensuring $\mathbf{A}_{l}$ is orthonormal\;
    Compute $I_{l}  = f(\mathbf{A}_{l})$\;
    \eIf{$I_{l} > I_{\text{cur}}$}{
      \KwRet{$\mathbf{A}_{l}$} \;
      }{
      $c = c + 1$\;
      }
    $l = l + 1$\;
  }
  \caption{random search}
  \label{random-search}
\end{algorithm}

@cook1995grand explained the use of a gradient ascent optimisation with an assumption that the index function is continuous and differentiable. Since some indices could be non-differentiable, the derivative is replaced by a pseudo-derivative of evaluating five randomly generated directions in a tiny nearby neighbourhood. Once the best direction is determined, the optimiser finds the best projection $\mathbf{A}_{**}$ on the geodesic between the current basis $\mathbf{A}_{\text{cur}}$ and the best directional basis $\mathbf{A}_{*}$. If the percentage change in the index value between $\mathbf{A}_{**}$ and $\mathbf{A}_{\text{cur}}$ is greater than a threshold value, $\mathbf{A}_{**}$ is outputted for the current iteration to construct the interpolation or the algorithm repeats the above steps until $l_{\max}$ is reached and the searching terminates. Algorithm \ref{search-geodesic} summarise the steps in geodesic search.

@@

Simulated annealing [@bertsimas1993simulated, @kirkpatrick1983optimization] is a non-derivative procedure based on a non-increasing cooling scheme $T(l)$. Given an initial $T_0$, the temperature at iteration $l$ is defined as $T(l) = \frac{T_0}{\log(l + 1)}$. The simulated annealing algorithm works as follows. Given a neighbourhood parameter $\alpha$ and a randomly generated orthonormal basis $A_{rand}$, a candidate basis is sampled by $\mathbf{A}_{t+1} = (1 - \alpha)\mathbf{A}_{t} + \alpha \mathbf{A}_{rand}$. If the index value of the candidate basis is larger than the one of the current basis, the candidate basis is outputted. While if it is smaller, the candidate has another chance of being accepted with probability $P= \min\left\{\exp\left[\frac{I_{cur} - I_{l}}{T(l)}\right],1\right\}$ where $I_{\cdot}$ denotes the index value of a given basis. The algorithm can be summarised as in Algorithm \ref{simulated_annealing}.

\begin{algorithm}
\SetAlgoLined
\SetKwInOut{Input}{input}
  \SetKwInOut{Output}{output}
    \Input{The current projection basis: $\mathbf{A}_{\text{cur}}$; The index function: $f$; Initial termperature: $T_0$}
    \Output{The global object; The target basis: $\mathbf{A}_{l}$}
  initialisation\;
  Set $l = 1$\;
  \While{$l < l_{\max}$}{
    Generate $\mathbf{A}_{l} = (1 - \alpha)\mathbf{A}_{\text{cur}} + \alpha \mathbf{A}_{rand}$ ensuring $\mathbf{A}_{l}$ is orthonormal \;
    Compute $I_{l} = f(\mathbf{A}_{l})$ and $T(l) = \frac{T_0}{\log(l + 1)}$\;
      \eIf{$I_{l} > I_{\text{cur}}$}{
        \KwRet{$\mathbf{A}_{l}$} \;
      }{
        Compute $P= \min\left\{\exp\left[\frac{I_{\text{cur}} -I_{l}}{T(l)}\right],1\right\}$\;
        Draw $D$ from a uniform distribution: $D \sim \text{Unif(0, 1)}$\;
        \If{$P > D$}{
           \KwRet{$\mathbf{A}_{l}$} \;
        }
      }
    $l = l + 1$\;
  }
  \caption{simulated annealing}
  \label{simulated_annealing}
\end{algorithm}

Notice that the three algorithms above constitute one search. The outputted basis is then sent to construct the interpolation and one such search-and-interpolate step is one round.

<!-- Below listed several features characterise the optimisation procedures needed in projection pursuit -->

<!-- - *Being able to handle non-differentiable PPI function*: The PPI function could be non-differentiable, thus derivative free methods are preferred.  -->

<!-- - *Being able to optimise with constraints*: The constraint comes from projection matrix being an orthonormal matrix.  -->

<!-- - *Being able to find both local and global maximum*: Although the primary interest is to find the global maximum, local maximum could also reveal structures in the data that are of our interest.  -->

@@

Random search methods has a black-box mechanism and focuses solely on finding the global maximum point, while the projection pursuit explores the data with an aim to discover the potential underlying structure and thus is interested in how the algorithm finds its maximum. This motivates us **to develop a visual diagnostic system for exploring the optimisation searching path**.

@@

Tidy data, as introduced in [@wickham2014tidy], has three criteria:  1) each observation forms a row, 2) each variable forms a column and 3) each type of observational unit forms a table. The adoption of tidy data makes it easier for data wrangling and visualisation. For our visual diagnostics, global object is a such data construction with key components in optimisation algorithms and is designed for making diagnostic plots easier. In the optimisation algorithms for projection pursuit, two key elements are the projection basis: $\mathbf{A}$ and the index value: $I$. State: $S$ labels the observation with detailed stage of searching or interpolation.

Multiple iterators are needed to index the data collected at different levels. $t$ is a unique identifier that prescribes the natural ordering of each observation; $j$ is the counter for each search-and-interpolate round, which remains the same for within one round and has an increment of one once a new round starts. $l$ is the counter for each search/interpolation allowing us to know how many basis has the algorithm searched before finding one to output.

@@

There are other parameters that are of our interest and we denote them as *$V_{p}$*. In projection pursuit, this includes $V_1 = \text{method}$, which tags the name of the algorithm used and $V_2 = \text{alpha}$, which is the neighbourhood parameter that controls the size in sampling candidate bases. The data structure can thus be shown as in Equation \ref{eq:data-structure}.

\begin{equation}
\left[
\begin{array}{c|ccc|cc|cc}
t & \mathbf{A} & I & S & j &  l  & V_{1} & V_{2}\\
\hline
1 & \mathbf{A}_1 & I_1 & S_1 & 1 & 1 & V_{11} & V_{21}\\
\hline
2 & \mathbf{A}_2 & I_2 & S_2 & 2 & 1  & V_{12}  & V_{22}\\
3 & \mathbf{A}_3 & I_3 & S_3 & 2 & 2  & V_{13}  & V_{23}\\
\vdots & \vdots &\vdots &\vdots  &\vdots & \vdots &\vdots  &\vdots\\
\vdots & \vdots & \vdots &\vdots & 2 & l_2 & \vdots  & \vdots\\
\hline
\vdots &\vdots & \vdots &\vdots & 2  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots & 2 & 2& \vdots &  \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & 2 & k_2 &\vdots  & \vdots\\
\hline
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
\hline
\vdots & \vdots & \vdots &\vdots  & J &  1 & \vdots & \vdots \\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T & \mathbf{A}_T & I_T &S_T  & J &  l_{J} & V_{1T}& V_{2T}\\
\hline
\vdots &\vdots & \vdots &\vdots & J  & 1& \vdots & \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots & \vdots  &\vdots \\
\vdots &\vdots &\vdots &\vdots & J & k_J &\vdots  & \vdots\\
\hline
\vdots& \vdots & \vdots & \vdots & J+1 & 1 & \vdots& \vdots\\
\vdots &\vdots &\vdots &\vdots &\vdots & \vdots &\vdots &\vdots \\
T^\prime & \mathbf{A}_{T^\prime} & I_{T^\prime} &S_{T^\prime}  & J+1 &  l_{J+1} & V_{1T^\prime}& V_{2T^\prime}\\
\end{array}
\right]
= 
\left[
\begin{array}{c}
\text{column name} \\
\hline
\text{search (start basis)} \\
\hline
\text{search} \\
\text{search} \\
\vdots \\
\text{search (outputed basis)} \\
\hline
\text{interpolate} \\
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\vdots \\
\hline
\text{search} \\
\vdots \\
\text{search (final basis)} \\
\hline
\text{interpolate} \\
\vdots \\
\text{interpolate} \\
\hline
\text{search (no output)} \\
\vdots \\
\text{search (no output)} \\
\end{array}
\right]
\label{eq:data-structure}
\end{equation}

where $T^{\prime} = T + k_{J}+ l_{J+1}$. Note that we deliberately denote the last round of search as $j = J+1$ and in that round there is no output/interpolation basis and the algorithm terminates. This notation allows us to denote the last complete search-and-interpolate round as round $J$ and hence the final basis is $A_T$ and highest index value found is $I_T$.

[outside the paper: I find the notation of current/target basis is confusing because the target basis in round $j$ becomes the current basis in round $j+1$. Also, when we start to have polish, the target basis may not be the current basis in the next round... The place where current/target is most appropriate is probably when describing the interpolation where the first one is always the current basis and the last is always the target basis. I think it is better to leave this language in the code]

A sketch of the global object for projection pursuit guided tour is presented in Figure \ref{fig:glb-obj}.

[I feel this sketch was initially useful but now since we have the data matrix and the printed output of the global object, it doesn't any additional information. I'm still keeping it here but we may need to remove it if it's not useful :(]

@@

We simulate some random variables of size 1000 with different structures. `x1`, `x8`, `x9` and `x10` are simulated from normal distribution with zero mean and variance of one as in equation \ref{eq:sim-norm}. When using projection pursuit to explore the data structure based on its departure from normality, the entry in the projection basis for these variables should be close to zero in theory. `x2` to `x7` are mixture of normal distributions with different weights and locations. Equation \ref{eq:sim-x2} to \ref{eq:sim-x7} outlines the distribution where each variable is simulated from and Figure \ref{origin-data} shows the histogram of each variable except `x3`. All the variables are then scaled to have variance of one before running the projection pursuit.

@@

We form our first dataset using variable `x1`, `x2`, `x8`, `x9` and `x10` and run the guided tour with optimiser `search_better`. When the optimisation ends, the global object will be stored and printed (it can be turned off by supplying argument `print = FALSE`). Additional messages during the optimisation can be displayed by argument `verbose = TRUE`. Below shows the first ten rows of the global object. Notice that the tibble object allows the list-column `basis` to be printed out nicely with the dimension of the projection basis readily available.

[I notice that in the implementation the loop index for interpolation starts from 0 rather than 1. Need more investigation to see if it is easy to change it to 1 or we will have to change the definition of the data structure.]

@@

```{r glb-obj-1}
holes_1d_better %>% 
  dplyr::select(id, basis, index_val, info, tries, loop, method, alpha) %>% 
  head(10)
```

@@

# Visual diagnostic plots {#vis-diag-plots}

@@

- Layer 1: boxplot geom
  - data: group by $j$ and filter the observations in the group that have count greater than `cutoff = 15`.
  - x: $j$ is mapped to the x-axis
  - y: the statistical transformed index value: $Q_q(I^{\prime}_t)$ is mapped to the y-axis where $Q_q(x)$ finds the $0, 25, 50, 75, 100$th-quantile for the vector $x$ and $I^{\prime}_t$ is the index value of all the searching bases defined in Matrix \ref{eq:data-structure}.

@@

# Implementation: Ferrn package {#implementation}
